---
title: "Exploratory Data Analysis - Diamond Prices"
author: "Casandra Philipson"
date: "February 12, 2016"
output: html_document
---

### Load the dataset and figure out the basic structure.

```{r}
library(ggplot2)
```

```
data("diamonds")
head(diamonds)
dim(diamonds)
names(diamonds)

```
We find that the dataset has the following observations:  
carat, cut, color, clarity, depth, table, price, x, y, z  
  
There are 3 ordered factors in the dataset (cut, color, and clarity). Their ranks can be determined using:

```
factor(diamonds$cut)
factor(diamonds$color)
factor(diamonds$clarity)

```

Using factor() we find out the following scales:  
__Levels cut__: Fair < Good < Very Good < Premium < Ideal  
__Levels color__: D > E > F > G > H > I > J  
__Levels clarity__: I1 < SI2 < SI1 < VS2 < VS1 < VVS2 < VVS1 < IF  

# Create a histogram of the price of all diamonds in the dataset
```{r}
qplot(x = price, data = diamonds, binwidth = 100,
      color = I('#000000'), fill = I('#F79420'))

```

This dataset has a right-tailed distribution with a median = 2401 and mean = 3933.   

### How many diamonds fall into different price categories?

```{r}

diamonds$cost_250 <- NA
diamonds$cost_250 <- ifelse(diamonds$price < 250, 1, 0)
      
diamonds$cost_500 <- NA
diamonds$cost_500 <-ifelse(diamonds$price < 500, 1, 0)
      
diamonds$cost_15000 <- NA
diamonds$cost_15000 <-ifelse(diamonds$price >= 15000, 1, 0)   

```

There are `r sum(diamonds$cost_250)` diamonds valued under $250.  
There are `r sum(diamonds$cost_500)` diamonds valued under $500.  
There are `r sum(diamonds$cost_15000)` diamonds valued at $15000 or greater.  

### Explore the largest peak in the price histogram
Try to limit the x-axis, alter the bin width, and setting different breaks in the x-axis to see if you can get more info on this bracket.

```{r}
qplot(x = price, data = diamonds, binwidth = 30,
      color = I('#000000'), fill = I('#808000')) +
  scale_x_continuous(breaks = seq(0, 2000, 1000), lim = c(0,2000)) +
  scale_y_continuous(breaks = seq(0, 3500, 500))

library(dplyr)
diamond_subset <- filter(diamonds, price < 1500)
summary(diamond_subset[1:7])

```

It seems like most of the diamonds that cost under $1500 are around 0.35 carats!

```
remember you can save your figures using ggsave('priceHistogram.png')
```

### Plot data for diamond prices based on cut as separate histograms

```{r}
library(gridExtra)

qplot(x = price, data = diamonds, binwidth = 75) +
  facet_wrap(~cut)

diamonds %>% filter(price == max(price))
diamonds %>% filter(price == min(price))
by(diamonds$price, diamonds$cut, summary)

```
Interestingly the distribution of price looks similaar for diamonds of very good, premium, and ideal quality.  

We also see that the highest price diamond is premium cut. The lowest price diamond is a tie between premium and ideal cuts. Finally, the lowest median price is found in the ideal cut category.  

### Let's look at the distributions again, make y-axis flexible

```{r}

qplot(x = price, data = diamonds, bins = 30) +
  facet_wrap(~cut, scales = "free") +
  scale_x_log10()

```

### Create a histogram of price per carat and facet it by cut

```{r}

qplot(x = price/carat, data = diamonds) +
  facet_wrap(~cut, scales = "free")

```


### Investigate the price per carat of diamons across different colors using boxplots
Create a plot based on calculated price/carat separated by color. Investigate the IQR.

```{r}

qplot(x = color, y = price/carat, data = diamonds,
      geom = 'boxplot', color = color) +
    coord_cartesian(ylim = c(0, 10000))

IQR(subset(diamonds, color == 'J' )$price)
IQR(subset(diamonds, color == 'D' )$price)


```


### Investigate the weight of the diamonds (carat) using a frequency polygon.  
Vary the different bin widths to find which carat size has a count greater than 2000.

```{r}

qplot(x = carat, data = diamonds, 
      binwidth = 0.01,
      geom = 'freqpoly',
      xlab = 'Carat Size',
      ylab = 'Number of Diamonds',
      color = carat) +
  coord_cartesian(ylim = c(0, 2500))

table(diamonds$carat)[table(diamonds$carat)>2000]

```
