---
title: "Diamonds-P2"
author: "Casandra Philipson"
date: "February 19, 2016"
output: html_document
---

This code is for the UDACITY "Data Analysis with R" Lesson 4 problem set. The built in diamonds data set will be used.

### Create a scatter plot of price versus x

```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(gridExtra))

ggplot(diamonds, aes(x = x, y = price)) +
  geom_point(alpha = 1/20) +
  coord_cartesian(xlim = c(3, 10))

```

Observations: Overall there appears to be a positive exponential correlation between price and x. The x-axis has been limited to remove a few of the outliers from the figure. There are a few interesting clusters at x ~ 7.5 and x ~ 8.25.

### What is the correlation between price and the variables x, y, and z?

```{r}

cor_x <- cor.test(diamonds$price, diamonds$x)
cor_y <- cor.test(diamonds$price, diamonds$y)
cor_z <- cor.test(diamonds$price, diamonds$z)

```

The correlation between price and the x/y/z variables are all close to 0.86

### Create a scatter plot of price versus depth

```{r}

ggplot(diamonds, aes(x = depth, y = price)) +
  geom_point(alpha = 1/100) +
  coord_cartesian(xlim = c(55, 70)) +
  scale_x_discrete(breaks = seq(55, 70, 2))


cor.test(diamonds$price, diamonds$depth)
```

### Create a scatter plot of price versus carat

```{r}

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/100, position = position_jitter(h = 0)) +
  coord_cartesian(xlim = c(0, quantile(diamonds$carat, 0.99)),
                  ylim = c(0, quantile(diamonds$price, 0.99)))
    
cor.test(diamonds$price, diamonds$carat)
```


### Create a scatter plot of price versus volume (x*y*z)

```{r}

diamonds$volume <- diamonds$x * diamonds$y * diamonds$z

#Are there any diamonds with a volume == 0?
tally(group_by(diamonds,volume == 0))

#What is the correlation of price nd volume? Exclude diamonds that have volume of 0 or are greater than or equal to 800.
with(subset(diamonds, volume > 0 & volume <= 800), cor.test(price, volume))

#Plot this subset, add transparency, add linear plot
ggplot(aes(x = volume, y = price), 
       data = subset(diamonds, volume > 0 & volume <= 800)) +
  coord_cartesian(ylim = c(0,20000)) +
  geom_point(alpha = 1/50) +
  geom_smooth(method = 'lm', color = 'red')

```

###Make two new data frames called diamonds_by_clarity and diamonds_by_color, then plot data 

The data frame should contain the following variables (in order):  
* mean_price  
* median_price  
* min_price  
* max_price  
* n  

```{r}
#create new data frame
diamonds_by_clarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(
            mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n())

diamonds_by_color <- diamonds %>%
  group_by(color) %>%
  summarise(
            mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n())

bar_p1 <- ggplot(diamonds_by_clarity, aes(x = factor(clarity), y = mean_price)) +
  geom_bar(stat = "identity", fill = "darkgreen")

bar_p2 <- ggplot(diamonds_by_color, aes(x = factor(color), y = mean_price)) +
  geom_bar(stat = "identity")

grid.arrange(bar_p1, bar_p2, ncol = 1)
  
```

Mean price tends to decrease as clarity improves. The same can be said for color.

### Working on a stacked bar chart for fun
Note: facet_wrap is a better alternative to stacked charts!  

```{r}
ggplot(diamonds, aes(clarity, fill=cut)) + geom_bar()
```